Toward single particle reconstruction without particle picking

### 2018-07-26 TEX/draft3.tex
כובל ופ רקבםהקרט

\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{verbatim}
\usepackage{float}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
%\usepackage{multirow}
\usepackage{color}
\usepackage{url}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{caption}
\usepackage{mathtools} 
\usepackage[margin=1.2in]{geometry}

@@

\paragraph{Particle picking.}
All contemporary methods in the field split the reconstruction procedure into several subroutines. 
The first step in the pipeline is  known as \emph{particle picking}, in which one identifies the locations of the 2-D tomographic projections of the samples embedded within the noisy micrograph. The output of an ideal particle picker is a series of 2-D images, each containing one centered tomographic projection associated with an unknown viewing direction and an in-plane rotation. This series of images 
is then used to estimate the structure.

Many automatic and semi-automatic methods for particle picking have been proposed, based on edge detection, template matching and deep learning; see for instance~\cite{harauz1989automatic,ogura2004automatic,zhu2016deep,frank1983automatic,scheres2015semi,heimowitz2018apple}. 
However, most of these procedures are prone to \emph{model bias}. For instance, in the popular framework of RELION~\cite{scheres2015semi}, the user manually marks hundreds of spots on the micrograph, believed to contain projections. 
Therefore, the algorithm's performance depends on the prior assumptions of the users about the particle's structure; the same holds true for deep learning based approaches which require constructing labeled sets of data.
Other methods use disks or differences of Gaussians as templates~\cite{langlois2014automated,voss2009dog}.
Nowadays, it is also still popular to pick particles manually. This method, while it exploits the researcher's experience, is both tedious and subject to model bias.

The significance of  model bias in cryo--EM was stressed in~\cite{shatsky2009method,henderson2013avoiding} and demonstrated by the ``Einstein from noise'' example.
In this experiment, the image of Einstein is correlated with multiple images of pure white noise. The noisy images are aligned to Einstein's image using cross-correlation and then averaged. 
In~\cite{shatsky2009method}, it was shown that with merely  1000 such noisy images, the averaged image is remarkably similar to the template image of Einstein rather than to pure noise.
%In Figure~\ref{fig:EinsteinFromNoise} we conducted this experiment with different number of noise images. With merely 1000 noise images, Einstein's face is clearly observed in the averaged image. With $10,000$  image, the output is remarkably similar to the template image of Einstein rather than to pure noise. Crucially, the more noisy images we use, the more severe the model bias towards the template. 
In the context of cryo--EM, this example demonstrates how our prior assumptions about the particle may influence the reconstructed structure.
This is true in general for any method which is based on template matching.
\TODO{TB: Do we have a nice experiment to exemplify the phenomenon?}

@@

\paragraph{Simplified model for SPR.} In this paper, we present a preliminary study that aims to show that estimating a particle 
directly from the micrograph, without identifying the location of individual particle images in the micrograph, might be possible.

@@

Figure~\ref{fig:micro_example} shows a simple example of our problem with  one signal ($K=1$). 
Each panel demonstrates an image of size $250\times 250$, containing 4 occurrences of an Einstein image of size $50\times 50$. 
The left panel shows the data without noise. In the middle panel,  i.i.d.\ Gaussian noise with standard deviation $\sigma=0.5$  was added.  At this noise level, Einstein's locations are still easily identified. In this scenario it is likely that standard detection algorithms can locate the copies of the signal within the micrograph. In the right panel, we show the same data swamped in  
i.i.d.\ Gaussian noise with standard deviation  $\sigma=3$.
Clearly, identifying the locations of Einstein in this micrograph is much more challenging. In fact, it can be shown that in the low signal--to--noise (\SNR) regime, 
detection  of individual image occurrences is impossible---even if the true image is known---and therefore particle picking is impossible.  This phenomenon is explained in detail in Appendix~\ref{sec:theory}. \TODO{WL: I think this example should go in the previous section on particle picking. The main issue we're trying to overcome is when detection is impossible; the fact that we don't use a template seems to be an additional advantage of our method, but not the main selling point. So I think the impossibility of detection in sufficiently low SNR must be emphasized earlier.}

\begin{figure}[h!]
	\centering
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_clean}
		\caption{$\sigma = 0$}
	\end{subfigure}%
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_s05}
		\caption{$\sigma = 0.5$}
	\end{subfigure}
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_s3}
		\caption{$\sigma = 3$}
	\end{subfigure}
RR	\caption{\label{fig:micro_example} Example of data sequences (micrographs) of size $250\times 250$ with additive i.i.d.\ Gaussian noise with variance $\sigma^2$. Each micrograph contains four occurrences of a $50 \times 50$ image of Einstein.}	
\end{figure}

\paragraph{Autocorrelation analysis.}
We now return to the general problem of estimating $x_1,\dots,x_K$ from the noisy observation $y$.
In order to recover the signals $x_1,\dots,x_K$ from the data, we use autocorrelation analysis.
The method consists of two stages. First, we estimate a mixture (i.e., linear combination) of the low--order autocorrelation functions of the signals from the data. These quantities can be estimated, to any desired accuracy, if individual occurrences are separated by the maximum signal size and each signal appears sufficiently many times in the data. There is no need to detect individual occurrences.
In the second stage, the signals are estimated from the mixed autocorrelations using a nonconvex least-squares (LS) algorithm or a phase retrieval algorithm; see Section~\ref{sec:methods} for details.
This method requires only one pass over the data and can recover the signals in any $\SNR$, if the signals appear enough times in the data. As a side note, we mention that expectation-maximization---a popular framework in SPR---is intractable for this problem (see Appendices~\ref{sec:related_literature} and~\ref{sec:theory}  for more details).

@@

In the first experiment, we estimated Einstein's image of size $50\times 50$ from multiple micrographs of size $4096\times 4096$ pixels, contaminated  with additive i.i.d.\ Gaussian noise with standard deviation $\sigma=3$ ($\SNR=1/20$). An illustrative example of a micrograph appears in the right panel of Figure~\ref{fig:micro_example}.
Each micrograph contains multiple occurrences of Einstein's image  in random locations, while keeping a separation of at least $49$ pixels between the images in each direction. From the micrographs, we estimated the signal's power spectrum, which is equivalent to the second-order autocorrelation. Then, in order to estimate the signal itself, we applied a standard phase retrieval algorithm called relaxed-reflect-reflect (RRR)~\cite{elser2017matrix}.

As an initial guess, we picked an image of the physicist Issac Newton; see Figure~\ref{fig:newton}. If the algorithm was prone to model bias, we would expect to get as an output an image that resembles Newton, similar to the ``Einstein from noise'' effect. However, the experiment exhibits the desired phenomenon: the more data we collect, the better the reconstruction quality.

Figure~\ref{fig:Einst_example} demonstrates several recovery results for different number of recorded micrographs. Figure~\ref{fig:error_per_micro} presents the normalized recovery error and the power spectrum estimation error as a function of the number of micrographs.
To measure the error, we use the root mean square error (RMSE) defined as 
\begin{equation}
\text{RMSE}  := \frac{\|x - \hat{x}\|_{\text{F}}}{\|x \|_{\text{F}}},
\end{equation} 
where $x$ and $\hat{x}$ are, respectively, the underlying and estimated image, and $\|\cdot\|_{\text{F}}$ stands for the Frobenius norm. 
As expected, the RMSE of estimating the power spectrum   decreases linearly with slope $-1/2$ in logarithmic scale. 
\TODO{TB: Few comments: 1. We have a movie in the supplementary material. 2. Does the error also have the right slope? We'll know once we'll gather enough micrographs 3. Need to improve the micrograph's generation code  }

\begin{figure}[h!]
	\centering
	\begin{subfigure}[h]{0.3\textwidth}
		\centering
		\includegraphics[scale=0.35]{Newton}
		\caption{\label{fig:newton}Newton  (model)}
	\end{subfigure}%
	\begin{subfigure}[h]{0.3\textwidth}
		\centering
		\includegraphics[scale=0.35]{reconstruction1}
		\caption{\small $P = 512$}
	\end{subfigure}%
	\begin{subfigure}[h]{0.3\textwidth}
		\centering
		\includegraphics[scale=0.35]{reconstruction10}
		\caption{$P = 512\times 10$}
	\end{subfigure}%

	\begin{subfigure}[h]{0.3\textwidth}
		\centering
		\includegraphics[scale=0.35]{reconstruction100}
		\caption{$P = 512\times 100$}
	\end{subfigure}%
	\begin{subfigure}[h]{0.3\textwidth}
	\centering
	\includegraphics[scale=0.35]{reconstruction200}
	\caption{$P = 512\times 200$}
	\end{subfigure}%
	\begin{subfigure}[h]{0.3\textwidth}
	\centering
	\includegraphics[scale=0.35]{reconstruction840}
	\caption{$P = 512\times 840$}
	\end{subfigure}%

	\caption{\label{fig:Einst_example} Recovery of Einstein's image. The RRR algorithm is initialized with the image of the physicist Issac Newton in the upper-left panel. We show estimations of Einstein using $P$ number of micrographs, each containing $700$ image occurrences on average with an $\SNR$ of $1/20$. The signal's estimation RMSE is  $0.5874, 0.3731, 0.2517, 0.2137, 0.1534$. The power spectrum estimation RMSE is $0.0381, 0.012, 0.0038, 0.0027, 0.0013$. }
\end{figure}

@@

Results from a similar experiment with three ($K=3$) one-dimensional signals are presented in Figure~\ref{fig:1Dheterosignals}.
 We define the ratio of the space occupied by the $i$th signal as 
\begin{equation}
\gamma_i = \frac{M_i L}{N},
\end{equation}
where $M_i$ is the number of times signal $x_i$  occurs, $L$ is the length of each signal and $N$ is the micrograph length.
These ratios $\gamma_i$ represent the {density} of the the signals in the data.
In the 1-D experiment, we do not assume knowledge of either the ratios $\gamma_i$ or the noise level $\sigma$.

In order to estimate the signals, we compute the first three autocorrelation functions of the data and then estimate the signals and their corresponding $\gamma_i$ using a nonconvex least-squares. 
As can be seen, given enough signal occurrences, we can estimate the signals accurately. 
The estimation quality of the triangle signal is poorer than the other two signals, a phenomenon that is explained using Proposition~\ref{prop:uniqueness}.
\TODO{TB: 1. Figure~\ref{fig:1Dheterosignals} will be replaced  with a ``progress figure'' (like we did for Einstien) for each signals and a plot of the recovery error for all three signals. 2. We may want to replace the triangle signal}
%We do not assume the knowledge of the noise level. 
%Crucially, starting from the point of $n\approx10^8$, the RMSE of the signal estimation decreases linearly with slope $-1/2$---the expected estimation rate of the autocorrelation functions---implying the stability of the recovery algorithm.

\begin{figure}[h]
\centering
\includegraphics[scale=.7]{Einstein_progress}
\caption{\label{fig:error_per_micro} The left panel shows the RMSE of estimating the power spectrum of the signal from the data as a function of the number of  collected micrographs. As expected, the curve is linear with slope $-1/2$ in logarithmic scale. The right panel shows the RMSE  (in logarithmic scale) of recovering Einstein's image. \TODO{TB: The program continues running so these figures will be extended to the right (more micrographs).}}	
\end{figure}

@@

\begin{figure}[t]
	\centering
	\includegraphics[width=\linewidth]{XP_1D_heterogeneous/hetero_example}
	\caption{An experiment for estimating three one-dimensional signals simultaneously with noise level $\sigma=3$. Red signals are the ground truth (targets) and the blue signals are our estimations. Individual RMSE of the estimates: $0.0239393 / 0.208925 / 0.0335956$. The estimated $\gamma$'s are $0.02574 / 0.01693 / 0.00818$ and the true ones $0.02561 / 0.01707 / 0.00853$. The $\SNR$ is $1/12$. \TODO{TB: 1. We will replace the figure with ``progress'' plots 2. Need to replace the triangle signal 3. Need to add error curves}}
	\label{fig:1Dheterosignals}
\end{figure}

@@

All current algorithmic pipelines for SPR using cryo--EM start with a particle picking stage to detect the 2-D tomographic projections of the samples in the micrograph.
The particle picking  algorithms tend to be prone to model bias and cannot work reliably on small particles in low $\SNR$ environment. 
Bypassing the particle picking stage and constructing a 3-D model directly from the data---without assuming prior knowledge about the particle to be estimated---can be used to reconstruct ab initio models to initialize a refinement algorithm.  Alternatively, it can be applied to generate templates for a particle picker which does not suffer from model bias.

In this paper, we examined a simplified model that, we believe, captures important features of the SPR problem. 
While the field is currently dominated by Bayesian methods such as EM, they are intractable for such problems. As an alternative, we propose to use an autocorrelation analysis technique. 
Methods based on autocorrelation analysis require only one pass over the data and therefore are usually computationally efficient. \TODO{TB: Do we want to mention Kam's method?}
%This property has been exploit before  for  ab inito modeling assuming ideal particle picking~\cite{kam1980reconstruction,levin20173d,singer2018mathematics}. 
That being said, the SPR model is far more complicated than the model presented here. In future research, we hope to bridge this gap.

Our results rely on two core assumptions that are not necessarily met by an SPR experiment.  
First, we modeled the background information as i.i.d.\ additive noise. In practice, the background information may be structured or signal dependent. 
Second, we assumed that the signal occurrences are all separated by the length of the signal; see~\eqref{eq:spacing}. 
This separation can be induced by careful experimental design \TODO{TB: ???}. 
If the signals are not separated, one can introduce additional parameters that model the distribution of the spacing  between signal occurrences and then compute explicitly the relation between the autocorrelation functions of the data and the signals. It is yet to be studied under what conditions on this distribution one can estimate the signals.

@@

Let $x_1,\ldots,x_K\in\RL$ be the target signals and let $y\in\RN$ be the observed data. 
The forward model can be posed as a mixture of \emph{blind deconvolution} problems between binary signals and the target signals $x_i$:
\begin{equation} \label{eq:model}
y = \sum_{i=1}^K x_i\ast s_i + \varepsilon,\quad \varepsilon\sim\mathcal{N}(0,\sigma^2 I).
\end{equation}
%We model the background information as i.i.d.\ Gaussian noise with zero mean and $\sigma^2$ variance. 
The nonzero values of each  $s_i\in\{0,1\}^N$ determine the position of $x_i$'s occurrences. 
We denote the set of these nonzero values by 
 $\mathcal{S}_i$ and its cardinality by $\vert \mathcal{S}_i\vert = M_i$. 
By assuming that all $\mathcal{S}_i$'s are disjoint,  we let $s = \sum_{i=1}^Ks_i$, $\mathcal{S} = \bigcup_{i=1}^{K} \mathcal{S}_i$ and  $\vert \mathcal{S}\vert :=M =  \sum_{i=1}^{K}M_i$.  %Neither the $M_i$'s nor $M$ are assumed to be known.
 Literature survey on blind deconvolution and related problems is given in Appendix~\ref{sec:related_literature}.

@@

\paragraph{Aperiodic autocorrelation functions.}
%\subsection{Aperiodic autocorrelation functions} \label{sec:aperiodic_ac}
For  
$z\in\RL$ and $k\geq 2$, the autocorrelation of order $k$ is defined for any integer shifts $\ell_1, \ldots, \ell_{k-1}$ by
\begin{align}
	a_z^k[\ell_1,\ldots,\ell_{k-1}]  & = \sum_{i=-\infty}^{+\infty} z[i]z[i+\ell_1]\ldots z[i+\ell_{k-1}],
	\label{eq:ac_general}
\end{align}
where the values of $z$ outside the indices $0, \ldots, L-1$ are taken to be zero.
Explicitly, the mean (which can be thought of as a first-order autocorrelation) and the second and third autocorrelations are given by
\begin{align} 
	a_z^1 & = \sum_{i=0}^{L-1} z[i], \nonumber\\
	a_z^2[\ell] & = \sum_{i = \max\{0, -\ell\}}^{L-1 + \min\{0, -\ell\}} z[i]z[i+\ell], \nonumber\\
	a_z^3[\ell_1,\ell_2] & = \sum_{i = \max\{0, -\ell_1, -\ell_2\}}^{L-1 + \min\{0, -\ell_1, -\ell_2\}} z[i]z[i+\ell_1]z[i+\ell_2]. \label{eq:ac_special}
\end{align}
Note that the autocorrelation functions are symmetric so that $a_z^2[\ell] = a_z^2[-\ell]$ and also $$a_z^3[\ell_1,\ell_2] = a_z^3[\ell_2,\ell_1]=a_z^3[-\ell_1,\ell_2-\ell_1].$$
%Additionally, if the moments of the signal depend only on the difference between the indices (Toeplitz structure), then they are equivalent to the autocorrelation functions.

@@

We first consider the problem of estimating the autocorrelations of a single signal from the data. 
The main principles carry through for $K>1$, as will be shown next.

%In order to estimate the autocorrelations of the signal, we first compute the first $L$ entries of the data's autocorrelations. 
For the purpose of the analysis, we consider  the asymptotic regime where $M$ grows proportionally to $N$. 
Specifically, we define the ratio of the measurement occupied by  the signal as
\begin{equation}
\gamma = \frac{M L}{N}.
\end{equation}
\TODO{WL: $\gamma$ should really be the limit of this ratio as $N$ grows to infinity; for any finite value of $N$ this ratio may be different. TB: The accurate definition is given in~\eqref{eq:gamma_limit}. We may want to leave it as is here.}
Under the spacing constraint~\eqref{eq:spacing}, we have $\gamma\leq\frac{L}{2L-1}\approx 1/2$.

@@

We will let the number of instances of the signal $M$ grow with $N$, and write $M=M_N$ to emphasize this. We assume $M_N$ grows proportionally with $N$, and define:
%
\begin{equation} \label{eq:gamma_limit}
\gamma = \lim_{N\to\infty} \frac{M_NL}{N}<1.
\end{equation}
%
We will assume that $M_N=\Omega(N)$, so that $\gamma>0$. In the sequel, we will suppress the explicit dependence of $M$ on $N$ for notational convenience.

>>>>

\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{verbatim}
\usepackage{float}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
%\usepackage{multirow}
\usepackage{color}
\usepackage{url}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{mathtools} 
\usepackage[margin=1.2in]{geometry}

@@

\paragraph{Particle picking.}
All contemporary methods in the field split the reconstruction procedure to several subroutines. 
The first step in the pipeline is the so-called \emph{particle picking}, in which one aims to detect the 2-D tomographic projections of the samples from the noisy micrograph. The output of ideal particle picker is a series of 2-D images, each containing one centered tomographic projection associated with an unknown viewing direction and an in-plane rotation. This series of images 
is then used to estimate the structure.

Many automatic and semi-automatic methods for particle picking have been proposed, based on edge detection, template matching and deep learning; see for instance~\cite{harauz1989automatic,ogura2004automatic,zhu2016deep,frank1983automatic,scheres2015semi,heimowitz2018apple}. 
However, most of these procedures are prone to \emph{model bias}. For instance, in the popular framework of RELION~\cite{scheres2015semi}, the user manually marks hundreds of spots on the micrograph, believed to contain projections. 
Therefore, the algorithm's performance depends on the prior assumptions of the users about the particle's structure; the same holds true for deep learning based approaches which require constructing labeled sets of data.
Other methods use disks or differences of Gaussians as templates~\cite{langlois2014automated,voss2009dog}.
Nowadays, it also still popular to pick particles manually. This method, while it exploits the researcher's experience, is both tedious and subject to model bias.

The significance of the model bias in cryo--EM was stressed in~\cite{shatsky2009method,henderson2013avoiding} and demonstrated by the ``Einstein from noise'' example.
In this experiment, the image of Einstein is correlated with multiple images of pure noise. The noisy images are aligned to Einstein's image using cross-correlation and then averaged. 
In~\cite{shatsky2009method}, it was shown that with merely  1000 noise images, the averaged image is remarkably similar to the template image of Einstein rather than to pure noise.
%In Figure~\ref{fig:EinsteinFromNoise} we conducted this experiment with different number of noise images. With merely 1000 noise images, Einstein's face is clearly observed in the averaged image. With $10,000$  image, the output is remarkably similar to the template image of Einstein rather than to pure noise. Crucially, the more noisy images we use, the more severe the model bias towards the template. 
In the context of cryo--EM, this example demonstrates how our prior assumptions about the particle may influence the reconstructed structure.
This true in general to any method which is based on template matching.
\TODO{Do we have a nice experiment to exemplify the phenomenon?}

@@

\paragraph{Simplified model for SPR} In this paper, we present a preliminary study that aims to show that estimating a particle 
directly from the micrograph, without identifying the location of individual particle images in the micrograph, might be possible.

@@

Figure~\ref{fig:micro_example} shows a simple example of our problem with  one signal ($K=1$). 
Each panel demonstrates an image of size $250\times 250$, containing 4 occurrences of an Einstein image of size $50\times 50$. 
The left panel shows the data without noise. In the middle panel,  i.i.d.\ Gaussian noise with standard deviation $\sigma=0.5$  was added, however,  Einstein is still easily identified. In this scenario it is likely that standard detection algorithms can locate the copies of the signal within the micrograph. In the right panel, we show the same data swamped in  
i.i.d.\ Gaussian noise with standard deviation  $\sigma=3$.
Clearly, identifying the locations of Einstein in this micrograph is much more challenging. In fact, it can be shown that in the low signal--to--noise (\SNR) regime, 
detection  of individual image occurrences is impossible---even if the true image is known---and therefore particle picking is impossible.  This phenomenon is explained in detail in Appendix~\ref{sec:theory}.

\begin{figure}[h!]
	\centering
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_clean}
		\caption{$\sigma = 0$}
	\end{subfigure}%
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_s05}
		\caption{$\sigma = 0.5$}
	\end{subfigure}
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_s3}
		\caption{$\sigma = 3$}
	\end{subfigure}
	\caption{\label{fig:micro_example} Example of data sequences (micrographs) of size $250\times 250$ with additive i.i.d.\ Gaussian noise with variance $\sigma^2$. Each micrograph contains four occurrences of a $50 \times 50$ image of Einstein.}	
\end{figure}

\paragraph{Autocorrelation analysis.}
We now return to the general problem of estimating $x_1,\dots,x_K$ from the noisy observation $y$.
In order to recover the signals $x_1,\dots,x_K$ from the data, we use autocorrelation analysis.
In a nutshell, the method consists of two stages. First, we estimate a mixture (i.e., linear combination) of the low--order autocorrelation functions of the signals from the data. These quantities can be estimated, to any desired accuracy, if individual occurrences are separated by the maximum signal size and each signal appears sufficiently many times in the data. There is no need to detect individual occurrences.
In the second stage, the signals are estimated from the mixed autocorrelations using a nonconvex least-squares (LS) algorithm or a phase retrieval algorithm; see Section~\ref{sec:methods} for details.
This method requires only one pass over the data and can recover the signals in any $\SNR$, if the signals appear enough times in the data. As a side note, we mention that expectation-maximization---a popular framework in SPR---is intractable for this problem (see Appendix~\ref{sec:theory}  for more details).

@@

In the first experiment, we estimated Einstein's image of size $50\times 50$ from multiple micrographs of size $4096\times 4096$ pixels, contaminated  with additive i.i.d.\ Gaussian noise with standard deviation $\sigma=3$ ($\SNR=1/20$). An illustrative example of a micrograph appears in the right panel of Figure~\ref{fig:micro_example}.
Each micrograph contains multiple occurrences of Einstein's image  in random locations, while keeping a separation of  $49$ pixels between the images in each direction. From the micrographs, we estimated the signal's power spectrum, which is equivalent to the second-order autocorrelation. Then, in order to estimate the signal itself, we applied a standard phase retrieval algorithm called relaxed-reflect-reflect (RRR).

As an initial guess, we picked an image of the physicist Issac Newton; see Figure~\ref{fig:newton}. If the algorithm was prone to model bias, we would expect to get as an output an image that resembles Newton, similarly to the ``Einstein from noise'' effect of Figure~\ref{fig:EinsteinFromNoise}. However, the experiment exhibits the desired phenomenon: the more data we collect, the better the reconstruction quality.

Figure~\ref{fig:Einst_example} demonstrates several recovery results for different number of recorded micrographs. Figure~\ref{fig:error_per_micro} presents the normalized recovery error and the power spectrum estimation error as a function of the number of micrographs.
To measure the error, we use the root mean square error (RMSE) defined as 
\begin{equation}
\text{RMSE}  := \frac{\|x - \hat{x}\|_{\text{F}}}{\|x \|_{\text{F}}},
\end{equation} 
where $x$ and $\hat{x}$ are, respectively, the underlying and estimated image, and $\|\cdot\|_{\text{F}}$ stands for the Frobenius norm. 
As expected, the RMSE of estimating the power spectrum   decreases linearly with slope $-1/2$ in logarithmic scale. 
\TODO{Few comments: 1. We have a movie in the supplementary material. 2. Does the error also have the right slope? 3. We can put more images to exemplify the progress 4. Need to improve the micrograph's generation code  }

\begin{figure}[h!]
	\centering
	\begin{subfigure}[h]{0.25\textwidth}
		\centering
		\includegraphics[scale=0.3]{Newton}
		\caption{\label{fig:newton}Newton  (model)}
	\end{subfigure}%
	\begin{subfigure}[h]{0.25\textwidth}
		\centering
		\includegraphics[scale=0.3]{reconstruction2}
		\caption{\small $P =1024$}
	\end{subfigure}%
	\begin{subfigure}[h]{0.25\textwidth}
		\centering
		\includegraphics[scale=0.3]{reconstruction20}
		\caption{$P = 1024\times 10$}
	\end{subfigure}%
	\begin{subfigure}[h]{0.25\textwidth}
		\centering
		\includegraphics[scale=0.3]{reconstruction200}
		\caption{$P = 1024\times 100$}
	\end{subfigure}%
	\caption{\label{fig:Einst_example} Recovery of Einstein's image. The RRR algorithm is initialized with the image of the physicist Issac Newton in the left panel. We show estimations of Einstein using $P=1024,1024\times 10$ and $1024\times 100$ number of micrographs, each contains $700$ image occurrences on average and the $\SNR$ is $1/20$. The signal's estimation RMSE is $0.5514$, $0.3494$ and $0.2112$, respectively. The power spectrum estimation RMSE is, respectively, $0.0269 $, $0.0083$ and $0.0027$. \TODO{Can add more images to exemplify the progress}}	
\end{figure}

@@

Results from a similar experiment with three ($K=3$) one-dimensional signals are presented in Figure~\ref{fig:1Dheterosignals}.
 We define the ratio of the space occupied by the $i$th signal as 
\begin{equation}
\gamma_i = \frac{M_i L}{N},
\end{equation}
where $M_i$ is the number of signal's  occurrences, $L$ is the length of the signal and $N$ is the micrograph length.
These ratios present the {density} of the the signals in the data.
In the experiment, we do not assume to know these ratios, neither the noise level $\sigma$.

In order to estimate the signals, we computed the first three autocorrelation functions of the data and then estimated the signals and their corresponding $\gamma_i$ using a nonconvex least-squares. 
As can be seen, given enough signal occurrences, we can estimate accurately the signals. 
The estimation quality of the triangle signal is poorer than the other two signals, a phenomenon that is explained using Proposition~\ref{prop:uniqueness}.
\TODO{1. We will replace the figure with a ``progress figure'' (like we did for Einstien) for each signals and a plot of the recovery error for all three signals. 2. We may want to replace the triangle signal}
%We do not assume the knowledge of the noise level. 
%Crucially, starting from the point of $n\approx10^8$, the RMSE of the signal estimation decreases linearly with slope $-1/2$---the expected estimation rate of the autocorrelation functions---implying the stability of the recovery algorithm.

\begin{figure}[h]
\centering
\includegraphics[scale=.7]{Einstein_progress}
\caption{\label{fig:error_per_micro} The left panel shows the RMSE of estimating the power spectrum of the signal from the data as a function of the number of  collected micrographs. As expected, the curve is linear with slope $-1/2$ in logarithmic scale. The right panel shows the RMSE  (in logarithmic scale) of recovering Einstein's image. \TODO{The program still runs so these figures will be extended to the right (more micrographs).}}	
\end{figure}

@@

\begin{figure}[t]
	\centering
	\includegraphics[width=\linewidth]{XP_1D_heterogeneous/hetero_example}
	\caption{An experiment for estimate three one-dimensional signals simultaneously with noise level $\sigma=3$. Red signals are the ground truth (targets) and the blue signals are our estimations. Individual RMSE of the estimates: $0.0239393 / 0.208925 / 0.0335956$. The estimated $\gamma$'s are $0.02574 / 0.01693 / 0.00818$ and the true ones $0.02561 / 0.01707 / 0.00853$. The $\SNR$ is $1/12$. \TODO{1. to replace with a ``progress'' plot 2. replace the triangle signal 3. add error progress figure}}
	\label{fig:1Dheterosignals}
\end{figure}

@@

All current algorithmic pipeline for SPR using cryo--EM start with a particle picking algorithm which is prone to model bias. 
Bypassing the particle picking stage and constructing a 3-D model directly from the data---without assuming prior knowledge on the particle to be estimated---can be used to reconstruct ab initio models to initialize a refinement algorithm.  Alternatively, it can be applied to generate templates for a particle picker which does not suffer from model bias.

In this paper, we examined a simplified model that, we believe, captures important features of the SPR problem. 
While the field is currently dominated by Bayesian methods such as EM, they are intractable for such problems. As an alternative, we propose to use autocorrelation analysis technique that shares some common lines (did you get the wordplay?)  with Kam's method for ab inito modeling~\cite{kam1980reconstruction,levin20173d,singer2018mathematics}. 
That being said, the SPR model is far more complicated than the model presented here. In a future research, we hope to bridge this gap.

Our results rely on two core assumptions that are not necessarily met by an SPR experiment.  
First, we modeled the background information as i.i.d.\ additive noise. In practice, the background information may be structured or signal dependent. 
Second, we assumed that the signal occurrences are all separated by the length of the signal, see~\eqref{eq:spacing}. 
This separation can be induced by careful experimental design \TODO{???}. 
If the signals are not separated, one can introduce a new variable that represents the distribution of the spacing between signal occurrences and then compute explicitly the relation between the autocorrelation functions of the data and the signals. It is yet to be studied under what conditions on this distribution one can estimate the signals.

@@

Let $x_1,\ldots,x_K\in\RL$ be the sought signals and let $y\in\RN$ be the data. 
The forward model can be posed as a mixture of \emph{blind deconvolution} problems between binary signals and the target signals $x_i$:
\begin{equation} \label{eq:model}
y = \sum_{i=1}^K x_i\ast s_i + \varepsilon,\quad \varepsilon\sim\mathcal{N}(0,\sigma^2 I).
\end{equation}
%We model the background information as i.i.d.\ Gaussian noise with zero mean and $\sigma^2$ variance. 
The nonzero values of each  $s_i\in\{0,1\}^N$ determine the position of $x_i$'s occurrences. 
We denote the set of these nonzero values by 
 $\mathcal{S}_i$ and its cardinality by $\vert \mathcal{S}_i\vert = M_i$. 
By assuming that all $\mathcal{S}_i$'s are disjoint,  we let $s = \sum_{i=1}^Ks_i$, $\mathcal{S} = \bigcup_{i=1}^{K} \mathcal{S}_i$ and  $\vert \mathcal{S}\vert :=M =  \sum_{i=1}^{K}M_i$.  %Neither the $M_i$'s nor $M$ are assumed to be known.
 Literature survey on blind deconvolution and related problems is given in Appendix~\ref{sec:related_literature}.

@@

\paragraph{Aperiodic autocorrelation functions}
%\subsection{Aperiodic autocorrelation functions} \label{sec:aperiodic_ac}
For  
$z\in\RL$ and $k\geq 2$, the autocorrelation of order $k$ is defined for any integer shifts $\ell_1, \ldots, \ell_{k-1}$ by
\begin{align}
	a_z^k[\ell_1,\ldots,\ell_{k-1}]  & = \sum_{i=-\infty}^{+\infty} z[i]z[i+\ell_1]\ldots z[i+\ell_{k-1}],
	\label{eq:ac_general}
\end{align}
where indexing of $z$ out of the bounds $0, \ldots, L-1$ is zero-padded, as usual.
Explicitly, the mean (which can be thought of as a first-order autocorrelation) and the second and third autocorrelations are given by
\begin{align} 
	a_z^1 & = \sum_{i=0}^{L-1} z[i], \nonumber\\
	a_z^2[\ell] & = \sum_{i = \max\{0, -\ell\}}^{L-1 + \min\{0, -\ell\}} z[i]z[i+\ell], \nonumber\\
	a_z^3[\ell_1,\ell_2] & = \sum_{i = \max\{0, -\ell_1, -\ell_2\}}^{L-1 + \min\{0, -\ell_1, -\ell_2\}} z[i]z[i+\ell_1]z[i+\ell_2]. \label{eq:ac_special}
\end{align}
Note that the autocorrelation functions are symmetric so that $a_z^2[\ell] = a_z^2[-\ell]$ and also $$a_z^3[\ell_1,\ell_2] = a_z^3[\ell_2,\ell_1]=a_z^3[-\ell_1,\ell_2-\ell_1].$$
%Additionally, if the moments of the signal depend only on the difference between the indices (Toeplitz structure), then they are equivalent to the autocorrelation functions.

@@

We first consider the problem of estimating the autocorrelations of a single signal from the data. 
The main principles carry through for $K>1$ as will be shown next.

%In order to estimate the autocorrelations of the signal, we first compute the first $L$ entries of the data's autocorrelations. 
For the purpose of the analysis, we consider  the asymptotic regime where $M,N\to\infty$, while preserving fixed ratio. 
Specifically, we define the ratio of the measurement occupied by  the signal as
\begin{equation}
\gamma = \frac{M L}{N}.
\end{equation}
Under the spacing constraint~\eqref{eq:spacing}, we have $\gamma\leq\frac{L}{2L-1}\approx 1/2$.

@@

We will let the number of instances of the signal $M$ grow with $N$, and write $M=M_N$ to emphasize this. We assume $M_N$ grows proportionally with $N$, and define:
%
\begin{equation}
\gamma = \lim_{N\to\infty} \frac{M_NL}{N}<1.
\end{equation}
%
We will assume that $M_N=\Omega(N)$, so that $\gamma>0$. In the sequel, we will suppress the explicit dependence of $M$ on $N$ for notational convenience.
